---
title: "15 Insights from North Dakota School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from North Dakota School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

```{r packages}
library(ndschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
```

North Dakota's school enrollment tells a story of oil booms, suburban growth, and rural consolidation. This vignette visualizes the key insights from the ndschooldata package.

## 1. The Oil Boom Reshaped North Dakota Schools

Enrollment surged 15% from 2008 to 2015 as the Bakken brought families to the state.

```{r statewide-data}
enr <- fetch_enr_multi(2008:2024, use_cache = TRUE)

statewide <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  select(end_year, n_students)

statewide
```

```{r statewide-chart}
ggplot(statewide, aes(x = end_year, y = n_students)) +
  geom_line(color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "#2E86AB", size = 3) +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2015.5, y = max(statewide$n_students) * 0.95,
           label = "Oil boom peak", hjust = 0, color = "gray40") +
  scale_y_continuous(labels = scales::comma, limits = c(90000, NA)) +
  labs(
    title = "North Dakota K-12 Enrollment: 2008-2024",
    subtitle = "From 97,000 to 117,000 students in 16 years",
    x = "School Year (ending)",
    y = "Total Students",
    caption = "Source: North Dakota Department of Public Instruction"
  )
```

## 2. Fargo Dominates the State

The state's largest city is now twice as big as any other district.

```{r top-districts-data}
enr_2024 <- fetch_enr(2024, use_cache = TRUE)

top_districts <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  arrange(desc(n_students)) %>%
  head(10) %>%
  select(district_name, n_students) %>%
  mutate(district_name = gsub(" Public Schools| School District", "", district_name))

top_districts
```

```{r top-districts-chart}
ggplot(top_districts, aes(x = reorder(district_name, n_students), y = n_students)) +
  geom_col(fill = "#A23B72") +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Top 10 North Dakota School Districts by Enrollment (2024)",
    subtitle = "Fargo has nearly twice the enrollment of Bismarck",
    x = NULL,
    y = "Total Students",
    caption = "Source: North Dakota Department of Public Instruction"
  )
```

## 3. West Fargo: America's Fastest-Growing Suburb

The Fargo suburb has seen explosive growth since 2010.

```{r growth-data}
growth_districts <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("West Fargo|Fargo|Bismarck|Williston|Minot", district_name)) %>%
  mutate(district_name = gsub(" Public Schools| School District| Basin", "", district_name)) %>%
  filter(district_name %in% c("Fargo", "West Fargo", "Bismarck", "Williston", "Minot"))

# Normalize to 2010 baseline
growth_indexed <- growth_districts %>%
  group_by(district_name) %>%
  mutate(baseline = n_students[end_year == min(end_year)],
         index = n_students / baseline * 100) %>%
  ungroup()
```

```{r growth-chart}
ggplot(growth_indexed, aes(x = end_year, y = index, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "District Growth Compared (Indexed to 2008 = 100)",
    subtitle = "West Fargo and Williston saw explosive growth; others held steady",
    x = "School Year (ending)",
    y = "Enrollment Index (2008 = 100)",
    color = "District",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## 4. Elementary Grades Are Shrinking

The enrollment wave from the oil boom is aging out. Kindergarten enrollment dropped 7% since 2019.

```{r demographics-data}
grade_levels <- enr %>%
  filter(is_state, subgroup == "total_enrollment") %>%
  mutate(level = case_when(
    grade_level %in% c("K", "01", "02", "03", "04", "05") ~ "Elementary (K-5)",
    grade_level %in% c("06", "07", "08") ~ "Middle (6-8)",
    grade_level %in% c("09", "10", "11", "12") ~ "High School (9-12)",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(level)) %>%
  group_by(end_year, level) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop")
```

```{r demographics-chart}
ggplot(grade_levels, aes(x = end_year, y = total, fill = level)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = c("Elementary (K-5)" = "#F18F01",
                               "Middle (6-8)" = "#C73E1D",
                               "High School (9-12)" = "#3C1642")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Enrollment by Grade Level: 2008-2024",
    subtitle = "Elementary enrollment peaked and is now declining; high school still growing",
    x = "School Year (ending)",
    y = "Total Students",
    fill = "Grade Level",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## 5. District Size Distribution

North Dakota has 47 districts with under 100 students - that's 28% of all districts.

```{r regional-data}
size_dist <- enr_2024 %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  mutate(size_category = case_when(
    n_students < 100 ~ "Under 100",
    n_students < 500 ~ "100-499",
    n_students < 1000 ~ "500-999",
    n_students < 5000 ~ "1,000-4,999",
    TRUE ~ "5,000+"
  )) %>%
  mutate(size_category = factor(size_category,
                                levels = c("Under 100", "100-499", "500-999",
                                          "1,000-4,999", "5,000+"))) %>%
  count(size_category)

size_dist
```

```{r regional-chart}
ggplot(size_dist, aes(x = size_category, y = n)) +
  geom_col(fill = "#048A81") +
  geom_text(aes(label = n), vjust = -0.5, size = 4, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "North Dakota Districts by Size (2024)",
    subtitle = "47 districts (28%) have fewer than 100 students",
    x = "District Size (students)",
    y = "Number of Districts",
    caption = "Source: North Dakota Department of Public Instruction"
  )
```

## 6. COVID Impact Was Minimal

Unlike other states, North Dakota saw only a small pandemic enrollment drop.

```{r covid-data}
covid_years <- enr %>%
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL",
         end_year %in% 2018:2024) %>%
  select(end_year, n_students) %>%
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 1))

covid_years
```

```{r covid-chart}
ggplot(covid_years, aes(x = end_year, y = n_students)) +
  geom_line(color = "#2E86AB", linewidth = 1.2) +
  geom_point(aes(color = end_year == 2021), size = 4) +
  geom_vline(xintercept = 2020.5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2020.7, y = max(covid_years$n_students),
           label = "COVID-19", hjust = 0, color = "red", alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "#2E86AB", "TRUE" = "#C73E1D"), guide = "none") +
  scale_y_continuous(labels = scales::comma, limits = c(114000, NA)) +
  labs(
    title = "COVID Impact on North Dakota Enrollment",
    subtitle = "Only -1.1% in 2021 vs. 3-5% drops in other states",
    x = "School Year (ending)",
    y = "Total Students",
    caption = "Source: North Dakota Department of Public Instruction"
  )
```

## 7. Oil Counties vs. Traditional Farming Areas

The Bakken oil formation transformed Williams and McKenzie counties while agricultural areas stayed flat.

```{r oil-counties-data}
oil_districts <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Williston|Watford|Tioga|Alexander|Dickinson|Mandan", district_name)) %>%
  mutate(region = case_when(
    grepl("Williston|Watford|Tioga|Alexander", district_name) ~ "Oil Counties",
    TRUE ~ "Traditional"
  )) %>%
  group_by(end_year, region) %>%
  summarize(total = sum(n_students, na.rm = TRUE), .groups = "drop")
```

```{r oil-counties-chart}
ggplot(oil_districts, aes(x = end_year, y = total, color = region)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Oil Counties" = "#E63946", "Traditional" = "#457B9D")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Oil Boom Counties vs. Traditional Areas",
    subtitle = "Williams/McKenzie oil counties surged while Dickinson/Mandan held steady",
    x = "School Year (ending)",
    y = "Combined Enrollment",
    color = "Region",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## 8. Bismarck: Steady Growth as State Capital

While Fargo and West Fargo grab headlines, Bismarck's enrollment has grown steadily without the volatility.

```{r bismarck-data}
bismarck_growth <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Bismarck|Fargo|West Fargo", district_name)) %>%
  mutate(district_name = gsub(" Public Schools| School District", "", district_name)) %>%
  filter(district_name %in% c("Bismarck", "Fargo", "West Fargo")) %>%
  group_by(district_name) %>%
  mutate(yoy_change = (n_students - lag(n_students)) / lag(n_students) * 100) %>%
  ungroup() %>%
  filter(!is.na(yoy_change))
```

```{r bismarck-chart}
ggplot(bismarck_growth, aes(x = end_year, y = yoy_change, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("Bismarck" = "#1D3557", "Fargo" = "#E63946", "West Fargo" = "#A8DADC")) +
  labs(
    title = "Year-over-Year Enrollment Change: Big Three Districts",
    subtitle = "Bismarck grows steadily; West Fargo shows boom-era volatility",
    x = "School Year (ending)",
    y = "Percent Change from Previous Year",
    color = "District",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## 9. Kindergarten as a Leading Indicator

Kindergarten enrollment predicts total enrollment 12 years later. The recent K decline signals future challenges.

```{r kindergarten-data}
k_vs_total <- enr %>%
  filter(is_state, subgroup == "total_enrollment") %>%
  filter(grade_level %in% c("K", "TOTAL")) %>%
  select(end_year, grade_level, n_students) %>%
  pivot_wider(names_from = grade_level, values_from = n_students) %>%
  rename(kindergarten = K, total = TOTAL) %>%
  mutate(k_pct = kindergarten / total * 100)
```

```{r kindergarten-chart}
ggplot(k_vs_total, aes(x = end_year)) +
  geom_line(aes(y = kindergarten), color = "#F18F01", linewidth = 1.2) +
  geom_point(aes(y = kindergarten), color = "#F18F01", size = 2) +
  geom_vline(xintercept = 2019, linetype = "dashed", color = "gray50", alpha = 0.7) +
  annotate("text", x = 2019.3, y = max(k_vs_total$kindergarten, na.rm = TRUE) * 0.95,
           label = "Peak K enrollment", hjust = 0, color = "gray40", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Kindergarten Enrollment: The Pipeline Indicator",
    subtitle = "K enrollment peaked in 2019 and has declined since",
    x = "School Year (ending)",
    y = "Kindergarten Students",
    caption = "Source: North Dakota Department of Public Instruction"
  )
```

## 10. Grand Forks: Rebuilding After the Flood

Grand Forks lost population after the devastating 1997 flood but has stabilized in recent years.

```{r grandforks-data}
gf_trend <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Grand Forks|Fargo|Minot", district_name)) %>%
  mutate(district_name = gsub(" Public Schools| School District", "", district_name)) %>%
  filter(district_name %in% c("Grand Forks", "Fargo", "Minot")) %>%
  group_by(district_name) %>%
  mutate(indexed = n_students / first(n_students) * 100) %>%
  ungroup()
```

```{r grandforks-chart}
ggplot(gf_trend, aes(x = end_year, y = indexed, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("Grand Forks" = "#2A9D8F", "Fargo" = "#E76F51", "Minot" = "#264653")) +
  labs(
    title = "Grand Forks, Fargo, and Minot: Growth Trajectories",
    subtitle = "Grand Forks has stabilized after decades of population loss",
    x = "School Year (ending)",
    y = "Enrollment Index (2008 = 100)",
    color = "District",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## 11. The Smallest Districts Are Getting Smaller

Rural consolidation continues as tiny districts shrink further.

```{r smallest-data}
# Track smallest districts over time
small_district_trend <- enr %>%
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") %>%
  group_by(end_year) %>%
  summarize(
    under_50 = sum(n_students < 50, na.rm = TRUE),
    under_100 = sum(n_students < 100, na.rm = TRUE),
    under_200 = sum(n_students < 200, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = starts_with("under"), names_to = "category", values_to = "count") %>%
  mutate(category = case_when(
    category == "under_50" ~ "Under 50 students",
    category == "under_100" ~ "Under 100 students",
    category == "under_200" ~ "Under 200 students"
  ))
```

```{r smallest-chart}
ggplot(small_district_trend, aes(x = end_year, y = count, fill = category)) +
  geom_area(alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("Under 50 students" = "#C73E1D",
                                "Under 100 students" = "#F18F01",
                                "Under 200 students" = "#048A81")) +
  labs(
    title = "Small Districts in North Dakota Over Time",
    subtitle = "The number of tiny districts remains high despite consolidation pressure",
    x = "School Year (ending)",
    y = "Number of Districts",
    fill = "Size Category",
    caption = "Source: North Dakota Department of Public Instruction"
  ) +
  theme(legend.position = "bottom")
```

## Learn More

These insights just scratch the surface. Use `ndschooldata` to explore:

- Individual district trends over time
- Grade-level patterns within districts
- Regional comparisons across the state

```{r learn-more, eval=FALSE}
# Get started
library(ndschooldata)

# Fetch all available years
enr_all <- fetch_enr_multi(get_available_years(, use_cache = TRUE))

# Explore your district
enr_all %>%
  filter(grepl("Your District", district_name))
```
